---
title: "Weights Components"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.width = 7,
                      fig.height = 5)
library(tidyverse)
library(PracTools)
```

# General knowledge

the goal of weighting is can provide approximately unbiased and consistent estimates of many different population quantities.

the steps:

* calculate the base weights (design weights)
* adjusted by unknown eligibility
* nonresponse adjustment
* calibration to population values

The product of four components is the final weight.

$$
w_i = d_{0i} *a_{1i} * a_{2i} * a_{3i}
$$

# Base weights

$d_{0i}$ is introduced to account for the probability that a case was selected for the sample.

$$
d_{0i} = \frac{1}{\pi_i}
$$

The base weight is designed to address the issue of unequal probabilities of selection across individuals or units, thereby restoring the 
representativeness of the sample to the target population. A higher weight indicates that the sampled unit represents a larger number of 
individuals in the population.

Sum of $d_{0i}$ should be equal to the total number of elements in the population

commonly used base weights include:

* SRS without replacement(epsem): $$ d_{0i} = \pi_i^{-1} = \frac{N}{n} $$
* stratified SRS without replacement: $$ d_{0hi} = \pi_{hi}^{-1} = \frac{N_{h}}{n_{h}}$$
* two-stage sampling leading to epsem: $$d_{0ij} = \pi_{ij}^{-1} = \frac{N}{m\bar{n}}$$
* multistage sampling 


# Adjusted by unknown eligibility



# Nonresponse adjustment

## weighting class adjustment

## propensity score adjustment

### propensity weighting

### propensity stratification

## classification algorithm

### chi-square automatic interaction detection (CHAID)

```{r}
# import data
data(smho98)
?smho98

# deal with data
# PSYREHAB is the variable of interest
# create a response indicator R
smho98$R = 1
smho98$R[is.na(smho98$PSYREHAB)==TRUE]=0
smho98$stratum6 = cut(smho98$STRATUM,c(0,2,4,8,10,13,16),
labels = 1:6)

## install CHAID
# Sys.setenv(PATH = paste("/opt/homebrew/bin", Sys.getenv("PATH"), sep = ":"))
# Sys.which("gfortran")

# install.packages("CHAID", repos="http://R-Forge.R-project.org", type = "source")
# install.packages("partykit")
# install.packages("CHAID", repos="http://R-Forge.R-project.org")
library(CHAID)
```

```{r}
# CHAID only allow categorical covariates
smho98$beds_c = 
    with(smho98, 
        cut(BEDS,
        breaks=quantile(BEDS,
        probs=seq(0,1, by=0.25),
        na.rm=TRUE),
        include.lowest=TRUE))
smho98$beds_c = factor(smho98$beds_c, labels=c("1","2","3","4") )

smho98$exp_c = 
    with(smho98, cut(EXPTOTAL,
        breaks=quantile(EXPTOTAL),
        include.lowest=TRUE))

smho98$exp_c = factor(smho98$exp_c, labels=c("1","2","3","4") )

smho98$SEENCNT_c = 
    with(smho98, cut(SEENCNT,
        breaks=quantile(SEENCNT,
        probs=seq(0,1, by=0.25)),
        include.lowest=TRUE))

smho98$SEENCNT_c = factor(smho98$SEENCNT_c, labels=c("1","2","3","4") )
```

```{r}
data.tree = 
    smho98[,c("stratum6","beds_c","exp_c","SEENCNT_c","R")]
data.tree = lapply(data.tree, factor)
```

```{r}
## CHAID model
set.seed(100)
chaid.tree = chaid(R ~ ., data=data.tree)
plot(chaid.tree)
```

```{r}
# ## CHAID fitted values
# chaid.fitted = chaid.tree$fitted
# chaid.tbl = table(chaid.fitted[,1],chaid.fitted[,2])
# chaid.prob = chaid.tbl[,2]/apply(chaid.tbl,1,sum)
# chaid.level = as.numeric(rownames(chaid.tbl))
# num.level = length(unique(chaid.level))

# ## assign an individual weight to each respondent
# smho98$chaid.wt = NA
# for (c in 1:num.level) {
# smho98$chaid.wt[chaid.fitted[,1]==chaid.level[c]]
# = 1/chaid.prob[c]
# }
# ## check weights calculation
# table(smho98$chaid.wt)
# sum(smho98$chaid.wt[which(smho98$R==1)])
```

### classification and regression trees (CART)

```{r}
## create data
data.tree = smho98[,c("stratum6","BEDS","EXPTOTAL",
"SEENCNT","R")]

## run CART model
library(rpart)
set.seed(100)
cart.tree = rpart(R ~ .,data=data.tree, method = "class")

# post(cart.tree, file = "tree.ps",
# title = "Classification Tree for Nonresponse Adjustments")

## CART fitted values and assign weights
cart.fitted = predict(cart.tree, type="prob")
data.tree$cart.wt = 1/cart.fitted[,2]
sum(data.tree$cart.wt[data.tree$R==1])
```